---
title: "Chapter 16: Large models — GPU acceleration using OpenCL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 16: Large models — GPU acceleration using OpenCL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GPU acceleration is an **optional** feature of the **glmbayes** modelling stack.
All modeling functions --- `glmb()`, `lmb()`, `rglmb()`, and related tools --- run
fully on the CPU regardless of whether OpenCL is available. No setup is needed for
standard use.

OpenCL is implemented in the compiled backend **glmbayesCore** (imported by
**glmbayes**). A CPU-only **glmbayes** install is normal when **glmbayesCore**
was built without OpenCL; enabling GPU acceleration means reinstalling
**glmbayesCore** from source with OpenCL available at compile time --- not
rebuilding **glmbayes** itself ( **glmbayes** has no C++ `configure` step).

Where GPU acceleration pays off is with **large models**: high-dimensional
predictor sets or large posterior sample sizes. Envelope construction and
evaluation in **glmbayesCore** --- gradient and log-posterior work at each
tangency grid point --- is embarrassingly parallel. Passing `use_opencl = TRUE`
to `glmb()` or `lmb()` can substantially reduce wall time. See
**`vignette("Chapter-A10", package = "glmbayesCore")`** for technical detail.

This chapter is the **glmbayes** user checklist. Backend-only installs should
also read **`vignette("Chapter-16", package = "glmbayesCore")`**.

# What you see when you load glmbayes

When **glmbayes** is attached interactively it checks whether GPU acceleration
appears feasible. If `has_opencl()` is `TRUE` --- meaning **glmbayesCore** was
compiled with OpenCL --- attach is silent.

If `has_opencl()` is `FALSE` **and** a GPU or OpenCL stack is detected, you
may see:

```
Note: glmbayes provides full CPU capability in this session
(e.g. glmb(), lmb(), Prior_Setup()). GPU acceleration is recommended
for bigger models and appears available. Reinstall glmbayesCore from
source with OpenCL at compile time to enable it; see vignette("Chapter-16",
"glmbayes") for install instructions.
```

On a machine with no GPU and no OpenCL stack, attach is silent.

To suppress the message:

```r
options(glmbayes.quiet_opencl_startup = TRUE)
```

# Enabling GPU acceleration: three steps

Work through these steps in order.

## Step 1: Check whether OpenCL is already enabled

```r
library(glmbayes)
has_opencl()
```

`has_opencl()` reports whether **glmbayesCore** in this session was built with
OpenCL. If `TRUE`, pass `use_opencl = TRUE` to `glmb()` or `lmb()` and you are
done. Otherwise continue.

## Step 2: Ensure backend dependencies are OpenCL-ready

**glmbayes** depends on **glmbayesCore**, which in turn depends on
**opencltools** and **nmathopencl**. For a GPU-enabled backend, **opencltools**
must be built with OpenCL support and the system must provide drivers, headers,
and an ICD loader.

Check **opencltools**:

```r
opencltools::has_opencl()
```

If `FALSE`, follow **`vignette("Chapter-01", package = "opencltools")`**, then
reinstall **opencltools** (and **nmathopencl**) from source:

```r
install.packages(
  c("opencltools", "nmathopencl"),
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

**What you need on your system** (summary; details in **opencltools** Chapter 01):

| Component | What it provides | Needed for |
|-----------|-----------------|------------|
| GPU driver | Exposes hardware to the OS | Runtime |
| OpenCL headers (`CL/cl.h`) | Required at compile time | Source build |
| OpenCL ICD loader | Dispatches to vendor runtime | Runtime |

## Step 3: Reinstall glmbayesCore from source

Reinstall **glmbayesCore** from source so its `configure` / `configure.win`
can enable OpenCL. **Reinstalling glmbayes alone does not compile OpenCL into
the backend.**

Prebuilt **CRAN** / **R-Universe** **glmbayesCore** binaries are often CPU-only.
Use `type = "source"` from R-Universe (or GitHub) after Step 2.

### Windows

Install **Rtools** and, if needed, **devtools**:

```r
install.packages("devtools")
install.packages(
  "glmbayesCore",
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

Development versions:

```r
devtools::install_github("knygren/glmbayesCore")
devtools::install_github("knygren/glmbayes")
```

### Linux / macOS

```r
install.packages(
  "glmbayesCore",
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

Then update **glmbayes** if needed (binary install is fine once **glmbayesCore**
has OpenCL):

```r
install.packages("glmbayes")
```

### After the install

```r
library(glmbayes)
has_opencl()
#> [1] TRUE
```

# Verifying the setup

```{r, eval=FALSE}
diagnose_glmbayes()
```

This combines **opencltools** host checks with compile-time status for
**glmbayesCore**. Example output:

```
=== glmbayes OpenCL Diagnostic Report ===
...
[OK] glmbayes was compiled with OpenCL support.
```

The final line refers to the **glmbayesCore** backend (wording is historical).

Common resolutions:

- **Driver / headers / runtime** → see **opencltools** Chapter 01.
- **Runtime probe failed** → `clinfo` outside R; ICD vendors on Linux.
- **Not compiled with OpenCL** → repeat Step 2, reinstall **glmbayesCore** from
  source (not **glmbayes** alone).

# Running a GPU-accelerated model

Pass `use_opencl = TRUE` to `glmb()` or `lmb()`:

```{r, eval=FALSE}
example(Cleveland)
```

Cleveland CPU vs OpenCL illustration (not executed during vignette build):

```{r, eval=FALSE}
library(glmbayes)
data("Cleveland")

ps <- Prior_Setup(
  hd ~ age + sex + cp + trestbps + chol +
    fbs + restecg + thalach + exang + oldpeak + slope + ca + thal,
  family = binomial(logit),
  data = Cleveland
)

t_cpu <- system.time({
  fit_cpu <- glmb(
    hd ~ age + sex + cp + trestbps + chol +
      fbs + restecg + thalach + exang + oldpeak + slope + ca + thal,
    family = binomial(link = "logit"),
    pfamily = dNormal(mu = ps$mu, Sigma = ps$Sigma),
    data = Cleveland,
    n = 20000,
    Gridtype = 2,
    use_parallel = TRUE,
    use_opencl = FALSE,
    verbose = FALSE
  )
})

t_gpu <- system.time({
  fit_gpu <- glmb(
    hd ~ age + sex + cp + trestbps + chol +
      fbs + restecg + thalach + exang + oldpeak + slope + ca + thal,
    family = binomial(link = "logit"),
    pfamily = dNormal(mu = ps$mu, Sigma = ps$Sigma),
    data = Cleveland,
    n = 20000,
    Gridtype = 2,
    use_parallel = TRUE,
    use_opencl = TRUE,
    verbose = FALSE
  )
})

t_cpu
t_gpu
```

```{r, echo=FALSE, out.width="100%"}
knitr::include_graphics(
  system.file("extdata", "cleveland_non_opencl_output_01.png", package = "glmbayes")
)
knitr::include_graphics(
  system.file("extdata", "cleveland_opencl_output_01.png", package = "glmbayes")
)
```

```{r, eval=FALSE}
summary(fit_gpu)
```

```{r, echo=FALSE, out.width="100%"}
knitr::include_graphics(
  system.file("extdata", "cleveland_summary_output_01.png", package = "glmbayes")
)
knitr::include_graphics(
  system.file("extdata", "cleveland_summary_output_02.png", package = "glmbayes")
)
```

# Appendix A: AMD GPUs on Linux (ROCm OpenCL)

On **Ubuntu 22.04 or 24.04 LTS**:

```sh
sudo apt-get install rocm-opencl-runtime
```

Supported ROCm GPUs include RX 7900 / 7800 / 7700 and Instinct MI200 / MI300.
See **`vignette("Chapter-01", package = "opencltools")`**.
