---
title: "Chapter 1: The glmbayes package-A General Overview"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 1: The glmbayes package-A General Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The *glmbayes* package produces iid samples for Bayesian Linear Models and Bayesian Genereralized Linear Models by providing Bayesian versions of the *lmb* and *glm* functions for classical models. Extensive efforts have been made to make the new functions, *lmb* and *glmb*, as similar as possible to their classical cousins. To that end, the functions add a single required argument (pfamily) and one optional argument (n) to those typically used by the classical functions. The Bayesian functions also generally inherit and/or provide implemented methods for nearly all of the methods available for the *lm* and *glm* functions so that the same type of outputs as from the classical functions can be readily produced.   

This vignette is designed to provide users an overview of the package and the various functionalities available. In the rest of this introduction, we briefly touch on the basics of loading the package and accessings various help resources and demos (which we recommend as a next step for users after they read this vignette).    The next section then gives an overview of all the core functions that most users are likely to use. This includes
the two aforementioned functions, functions used to specify and check prior distributions, and methods available for the two functions. Most users can likely stop after that section and can then turn to either the help pages or the next set of vignettes that go into a more detailed comparisons of how the methods are used for these Bayesian functions vs. for their classical cousins.

For advanced users who are interested in implementing Block-Gibbs sampling procedures or in doing simulation for other more complex models, we recommend continuing on to the third section in this vignette that covers the two functions *rlmb* and *rglmb* and are designed for implementation of precisely that type of sampling. The third section discusses how those functions can be called, references some examples/demos where they are used, and describes how to use the output from the functions to access similar output to that produced by the *lmb* and *glmb* functions.

For users who are interested in the details of how the simulation is performed there is the final section that contains a discussion of the various simulation functions that are used to sample from the posterior distributions.  This includes a discussion of three core function that simuate for the currently implemented prior families and a couple of functions related to what we call the Central Normal distribution (the normal distribution between a lower and an upper bound). It also includes a discussion of a set of functions that are used in order to utilize the enveloping functions needed to get iid samples for some of the posterior distributions. That material is rather technical and we refer the user to the Nygren and Nygren (2006) paper that forms the foundation for the simulation procedure. 



## Loading the package 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package curently requires the MASS package so it also needs to be installed and gets loaded as part of the loading of the glmbayes package (see below).

```{r setup,results = "hide"}
library(glmbayes)
```

## Accessing package information, general documentation, and demos

To access information for the package outside of the information contained in these vignettes, we suggest users review three sets of help pages. A call to help(glmbayes) yields a page with a detailed discussion of the package with an example at the bottom.   Calling help(package="glmbayes") brings the user to a list of Help pages for specific functions as well as links to a general description file, code demos, and these vignettes. Part of the objective of this vignette is to provide more organized layout of the functions than the general list of help pages provided on the documentation page.

A call to demo(package="glmbayes"), finally, brings up a list of available demos for the package. A useful start for users might be to go through the specific demos as they are set up to illustrate the functionality of the package.

```{r help,results = "hide"}
help(glmbayes)
```

```{r documentation,results = "hide"}
help(package="glmbayes")
```

```{r demos,results = "hide"}
demo(package="glmbayes")
```

# Using the *lmb* and *glmb* functions and their methods

As noted above, the *lmb* and *glmb* functions are the two core functions in this package. More detailed side-by-side comparisons of the use of methods for these functions and their classical cousins are provided in the next two vignettes. Here we provide a brief overview of supporting functions available to facilitate the use of the core functions. In the first subsection, we show how the core functions are called (and how this compares to calls for the classical functions). We then give a brief discussion of how priors are specified and some related functions designed to assist with that task as well as with the task of viewiw prior specifications after the modeling is completed.   

## The *lmb* and *glmb* functions

The code snippets below illustrates how model specification for these models can be done by simply providind an additional argument, a pfamily, that provides information on the type of prior distribution that has been selected and the parameter values that are desired. A separate vignette is provided to give guidance to users on how to select appropriate prior parameters (and some standardization steps that may be useful as part of that).

[NEED TO MODIFY Prior_Setup PRINT FUNCTION AND HOW IT HANDLES CASE WHERE X IS MISSING]
[MAY ALSO WANT TO HAVE IT INITIALIZE SHAPE AND RATE PARAMETERS]

*Calling lm vs. calling lmb*

Here is the call for the lmb function and how it compares to the call to the classical lm function. The recommended prior here is the dNormal_Gamma pfamily which requires four parameters (2 for he Normal component, and 2 for the Gamma component).

```{r lm and lmb,results = "hide"}
# Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
##Classical model-call
lm.D9 <- lm(weight ~ group)

# Bayesian Prior Setup and Call-Prior Should Generally be Customized
p_info=Prior_Setup_v2(weight~group)
mu1=p_info$mu    
Sigma1=p_info$Sigma

lmb.D9 <- lmb(weight ~ group,pfamily=dNormal_Gamma(mu1,Sigma1,shape=4,rate=0.1))
```

*Calling glm vs. calling glmb*

The call for the glmb function is quite similar and is here compared to the corresponding call for the glm function. For the binomial and Poisson families, the recommended prior here is the dNormal pfamily which requires 2 parameters (a prior mean vector and a prior Variance-Covariance matrix).

```{r glm and glmb,results = "hide"}
## Dobson (1990) Page 93: Randomized Controlled Trial :
counts <- c(18,17,15,20,10,20,25,13,12)
outcome <- gl(3,1,9)
treatment <- gl(3,3)
##Classical model-call
glm.D93 <- glm(counts ~ outcome + treatment, family = poisson())

# Bayesian Prior Setup and Call-Prior Should Generally be Customized
p_info2=Prior_Setup_v2(counts ~ outcome + treatment)
mu2=p_info2$mu    
Sigma2=p_info2$Sigma

glmb.D93<-glmb(counts ~ outcome + treatment, family = poisson(),pfamily=dNormal(mu=mu2,Sigma=Sigma2))

```
## Specifying Priors

### Prior Family Functions

The *pfamily* arguments described above are actually functions that provide needed information to the *lmb* and *glmb* functions.  Since they are are functions, they can also be called directly, as seen below, where they return lists with items related to the prior (see documentation of these pfamily functions for details). These functions are also discussed in more detail in a latter chapter. 

While the current package only provides 3 implemented pfamiles, this is actually not a severe limitations as users easily can write their own prior families and pass those to the *lmb* and *glmb* functions.  One of our exaxmple includes an illustration of this and this approach is discussed more extensively in one of our other vignettes. Below, we simply illustrate how the functions can be called directly.

NOTE: IMPLEMENT THE EXAMPLE DISCUSSED ABOVE AND WRITE A SEPARATE VIGNETTE TO ACCOMPANY THE EXAMPLE.

*Calling the dNormal prior family*

This prior is typically used as a prior for the Poisson and Binomial families when calling either the
*glmb* or *rglm* functions. Calling it directly provides a list with, among other items, the prior specification.  the documentation for the *pfamilies* provides additional information on what the returned list contains and what
required and optional arguments are.

```{r Calling dNormal,results = "hide"}
dNormal(mu=mu2,Sigma=Sigma2,dispersion=NULL)
```
*Calling the dNormal_Gamma prior family

This prior is typically used as a prior when calling either the *lmb* or *rglmb* functions or when using the *gaussian()* family in the *glm* and *rglmb* functions. Calling it directly 

```{r Calling dNormal_Gamma,results = "hide"}

dNormal_Gamma(mu=mu1,Sigma=Sigma1,shape=4,rate=0.1)
```

*Calling the dGamma prior family

This prior is typically used when giving a prior for the dispersion parameter in the *lmb*, *rlmb*, *glmb*, and *rglmb* functions (in the latter case for the *gaussian()* and *Gamma()* families). Typicall this would be done as
part of Block-Gibbs sampling where the regression coefficients are updated in a separate block.

```{r Calling dGamma,results = "hide"}
b=lm.D9$coefficients
dGamma(shape=4,rate=0.1,beta=b)
```


### Other Prior Related Functions 

Three additional functions relate to the prior specification and are useful when either specifying the prior or 
when extracting the information about the prior used by an existing model object.

*The Prior_Setup function

This function can be used to set up the prior arguments so they have the correct dimensions and so that the various dimensions have names that correspond to those in the model. In addition to initialized prior structrues, these functions also return the model frame and the model matrix as a reference (the latter when it was requested from the underlying model). It is worth noting that this function is intended as a tool for initializig the prior objects and not for setting them to the final values used in the estimation. How to populate the arguments with sensible values is discussed in a separate vignette.

NOTE: ADD A PRINT METHOD FOR THIS FUNCTION SO THAT THE OUTPUT CAN BE CONTROLLED

```{r Calling Prior_Setup,results = "hide"}
Prior_Setup_v2(weight ~ group)
```
*The Prior_Check function

The Prior Check function provides a utility for checking whether the data appears to be consistent with the prior. Essentially, consistency here means that the maximum likelihood estimate are within the credible intervals implied by the prior. If the data appears inconsistent with the data, it is possible that the data was generated from a different process than that implied by the model or that the prior was not well thought out.  Specifying a good prior is usually important and we dedicate a separate vignette to this topic.


```{r Calling Prior_Check,results = "hide"}
Prior_Check(glm.D93,mu2,Sigma2)
```


*The pfamily function

This function can be used to extract a pfamily object from an existing object and is useful post-modeling in order
to verify the prior used in the modeling process. It is also used by some of the methods during post-processing
to extract information about the prior.

```{r Calling pfamily,results = "hide"}
pfamily(glmb.D93)
```


## Method functions 

The *lm* and *glm* functions come with an extensive set of method functions that can be used to conduct supplemental analysis (typically during the post-modeling phase). Because the *lmb* and *glmb* functions are structured to return many of the same items as the classical functions and because they inherit methods from *glm* and *lm*, many of the methods available for the classical functions are directly available for the *glmb* and *lmb* functions. As some of the output does tend to be different (in particular, coefficients are returned as random draws as opposed to a single maximum likelihood estimate), we have implemented specific methods for the "glmb" and "lmb" classes of objects.  The current set of methods for all four classes of objects are displayed below. 

Two separate vignettes illustrate how calls to these methods can be used to replicate the output from the classical models. It is worth noting that a smaller subset of the classical methods are not yet implemented (although plans are in place to replicate some of those as well). Those are discussed in the separate vignette's as well. Here we provide lists of the methods available for classical *lm* and *glm* functions (many of which are inherited by *lmb* and *glmb*) as well as methods that have been developed specifically for *glmb* and *rglmb*.

[NOTE: AS NEEDED, ADD ADDITIONAL METHODS FOR lmb. THE LIST OF METHODS FOR glmb SHOULD BE FAIRLY COMPLETE]

*Methods for class="lm"* 
```{r lm methods}
methods(class="lm")
```

*Methods for class="glm"* 

```{r glm methods}
methods(class="glm")
```
*Methods for class="glmb"* 

While most of the methods for the *glm* class (when it has its own method) or the *lm* class (when it does not) work properly for objects of class *glmb*, some methods do not. For that reason, we have implemented the below methods specifically for the *glmb* class of objects.  Future enhancements to this package may enhance these methods.  There are also additional methods for *lm* and/or *glm* that do not currently have functioning methods
for the *glmb* class. This includes the *add1*, *drop1*, and *dropterm* functions (which should be implementable fairly easily) and the various influence.measure methods (influence, rstandard, rstudent, dfbeta, dfbetas, cooks.distance, and hatvalues) which may require more work. Plans do exist to add the former and perhaps the latter (which likely would look at influence on posterior modes and not on posterior means).   

```{r glmb methods}
methods(class="glmb")
```

NOTE: CHECK WHICH METHODS FOR lmb that are inherited and functions vs. those that may require additional development.

```{r lmb methods}
methods(class="lmb")
```


# The *rlmb* and *rglmb* functions and their summary functions

Because the functions *lmb* and *glmb* are set up to mirror their corresponding classical cousins, they include a great deal of overhead for pre and post processing that are unsuitable if the functions need to be called repetitively (say during Block-Gibbs sampling).  To faciliate Block-Gibbs sampling and other simulation, we provide more minimalistic interfaces to the simulation procedures in the form of the corresponding functions *rlmb* and *rglmb*. 

These functions are called internally by *lmb* and *glmb* in much the same way that *lm.fit* and *glm.fit* are called by the *lm* and *glm* functions so in a way they are Bayesian versions of those functions. We choose the r prefix for these functions to symbolize that they also represent random draws from probability distributions in much the same way that *rnorm*, *rgamma*, and other functions in the *stats* package do.

## Calling the *rlmb* and *rglmb* functions

Instead of passing a formula to these functions, the user instead passes the dependent and independent variables as a dependent variable vector *y* and as an independent variable design matrix *x* respectively. A useful approach that can help the user specify the design matrix involves sequential calls to *model.frame* and *model.matrix* as seen below.

NOTE 1: SHORTEN THIS CODE
NOTE 2: The SUMMARY FUNCTION IS CURRENTLY NOT WORKING FOR rlmb

*Calling the rlmb function*
```{r calling lmb,results = "hide"}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
## Classical call
lm.D9 <- lm(weight ~ group,x=TRUE,y=TRUE)
lm_summary=summary(lm.D9)
## Bayesian Call
n<-10000
y<-lm.D9$y
#x<-as.matrix(lm.D9$x)

### Set old regression and dispersion coefficients

b_old=lm.D9$coefficients
v_old=lm_summary$sigma^2

#### Set up for rglmb_dispersion

n0=0.1
shape=n0/2
rate=shape*v_old

rate/(shape-1)
rate/shape

#a1<-shape+n1/2
#b1<-rate+sum(SS)/2

#out<-1/rgamma(n,shape=a1,rate=b1)
## v0=sum(SS)/n1 ~ (2*rate+sum(SS))/(2*shape+n1)=[(2*rate+sum(SS))/2]/((2*shape+n1)/2) 
n1=length(y)
SS=v_old*n1

n1 # This is equal to 20


n0=2 # Prior observations
v_prior=v_old  # Prior point estimate for variance (the mean of (1/dispersion=1/v_prior))
wt0=(n0/n1)  

## set shape=0.01*(n1/2)
## set rate= 0.01*SS/2

shape=wt0*(n1/2)   ###  Shape is prior observations /2
rate=shape*v_prior  ### rate is essentiall prior SS - V in rmultireg should be this
rate/shape ## Should match v_prior (currently also v_old)

mu<-c(0,0)
mu=b_old  ### For testing purposes, set prior=b_old
P<-0.1*diag(2)
# Bayesian Prior Setup and Call-Prior Should Generally be Customized
p_info=Prior_Setup_v2(weight~group)
x=p_info$x
outtemp4=rlmb(n=1000,y=y,x=x,pfamily=dNormal_Gamma(mu=mu,Sigma=solve(P),shape=shape,rate=rate))

#summary(outtemp4)

```

*Calling the rglmb function*

```{r calling rglmb}
data(menarche2)
Age2=menarche2$Age-13

y=menarche2$Menarche/menarche2$Total
wt=menarche2$Total

## Use model.frame and model.matrix to derive x
#mf=model.frame(formula)
#x=model.matrix(formula,mf)

x<-matrix(as.numeric(1.0),nrow=length(Age2),ncol=2)
x[,2]=Age2

## Modify Prior_Setup so it can take a model matrix as well as an model object
mu<-matrix(as.numeric(0.0),nrow=2,ncol=1)
V1<-1*diag(as.numeric(2.0))
mu[2,1]=(log(0.9/0.1)-log(0.5/0.5))/3

# 2 standard deviations for prior estimate at age 13 between 0.1 and 0.9
## Specifies uncertainty around the point estimates

V1[1,1]<-((log(0.9/0.1)-log(0.5/0.5))/2)^2 
V1[2,2]=(3*mu[2,1]/2)^2  # Allows slope to be up to 3 times as large as point estimate 

out<-rglmb(n = 1000, y=y, x=x, pfamily=dNormal(mu=mu,Sigma=V1), weights = wt, 
           family = binomial(logit)) 

summary(out)
```


## Methods for The *rlmb*, *rglmb*, and their summary functions

To allow for the output from the rglmb and rlmb functions to benefit from the wide range of methods available 
for their *lmb* and *glmb* cousins, we implement a few additional methods for these functions that enable most of the methods to also work for these objects post modeling.

```{r rglmb methods}
methods(class="rglmb")
```

```{r summary.rglmb methods}
methods(class="summary.rglmb")
```


```{r rlmb methods}
methods(class="rlmb")
```

# Advanced Topics



## Simulation Functions

## Normal_ct Functions

## Envelope Related Functions

## Utility Functions 
