---
title: "Chapter 6: Reviewing Model Predictions, Deviance Residuals and Model Statistics"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 6: Reviewing Model Predictions, Deviance Residuals and Model Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---
references:
- id: nygren2006
  title: Likelihood Subgradient Densities
  author:
  - family: Nygren
    given: Kjell
  - family: Nygren
    given: Lan Ma
  container-title: Journal of the American Statistical Association
  volume: 101
  URL: 'https://doi.org/10.1198/016214506000000357'
  DOI: 10.1198/016214506000000357
  issue: 475
  publisher: Taylor and Francis, Ltd
  page: 1144-1156
  type: article-journal
  issued:
    year: 2006
    month: 9

- id: Venables2002
  title: Modern Applied Statistics with S.
  author:
  - family: Venables
    given: W. N.
  - family: Ripley
    given: B. D.
  URL: 'https://www.springer.com/gp/book/9780387954578'
  DOI: 10.1007/978-0-387-21706-2
  publisher: Springer-Verlag New York
  page: 1-498
  type: book
  issued:
    year: 2002
    
- id: Hamilton2004
  title: Lecture notes obstetrics and gynaecology
  author:
  - family: Hamilton-Fairley
    given: Diana
  URL: 'https://www.bookdepository.com/Lecture-Notes-Obstetrics-Gynaecology-Diana-Hamilton-Fairley/9781405120661'
  publisher: Blackwell Pub.
  type: book
  issued:
    year: 2004
    month: 11

- id: Evans2006
  title: Checking for Prior-Data Conflict
  author:
  - family: Evans
    given: Michael
  - family: Moshonov
    given: Hadas
  container-title: Bayesian Analysis
  volume: 1
  URL: 'http://dx.doi.org/10.1214/06-BA129'
  DOI: 10.1214/06-BA129
  issue: 4
  page: 893-914
  type: article-journal
  issued:
    year: 2006
    month: 

- id: Spiegelhalter2002
  title: Bayesian measures of model complexity and fit
  author:
  - family: Spiegelhalter
    given: David J.
  - family: Best
    given: Nicola G.
  - family: Carlin
    given: Bradley P.
  - family: Van Der Linde
    given: Angelika
  container-title: Journal of the Royal Statistical Society Series B (Statistical Methodology)
  volume: 64
  URL: 'https://doi.org/10.1111/1467-9868.00353'
  DOI: 10.1111/1467-9868.00353
  issue: 4
  page: 583-639
  type: article-journal
  issued:
    year: 2002
    month: 9

    
---

---
nocite: | 
  @nygren2006
...



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(glmbayes)
```


```{r menarche data,results = "hide",echo = FALSE}
## Load menarche data
data(menarche2)
head(menarche2, 5)

```

```{r Analysis Setup,echo = FALSE}
## Number of variables in model
Age=menarche2$Age
nvars=2
set.seed(333)

## Reference Ages for setting of priors and Age_Difference
ref_age1=13  # user can modify this
ref_age2=15  ## user can modify this

## Define variables used later in analysis
Age2=Age-ref_age1
Age_Diff=ref_age2-ref_age1

```


```{r Prior Info,echo = FALSE}

## Point estimates at reference ages
m1=0.5  
m2=0.9

## Lower bound of prior credible intervals for point estimates
m1_lower=0.3
m2_lower=0.7

## Assumed correlation between the two (on link scale)
m_corr=0.4
```


```{r Logit: set up link function info and initialize prior matrices,echo = FALSE}

## Set up link function and initialize prior mean and Variance-Covariance matrices
bi_logit <- binomial(link="logit")
mu1<-matrix(0,nrow=nvars,ncol=1)
rownames(mu1)=c("Intercept","Age2")
colnames(mu1)=c("Prior Mean")
V1<-1*diag(nvars)
rownames(V1)=c("Intercept","Age2")
colnames(V1)=c("Intercept","Age2")
```


```{r Logit:set prior means,results = "hide",echo = FALSE}
## Prior mean for intercept is set to point estimate 
## at reference age1 (on logit scale)
mu1[1,1]=bi_logit$linkfun(m1)

## Prior mean for slope is set to difference in point estimates
## on logit scale divided by Age_Diff

mu1[2,1]=(bi_logit$linkfun(m2) -bi_logit$linkfun(m1))/Age_Diff 
print(mu1)

```

```{r Logit:set prior Variance Covariance matrix,results = "hide",echo = FALSE}
## Implied standard deviations for point estimates on logit scale

sd_m1= (bi_logit$linkfun(m1) -bi_logit$linkfun(m1_lower))/1.96
sd_m2= (bi_logit$linkfun(m2) -bi_logit$linkfun(m2_lower))/1.96

## Implied Standard deviation for slope (using variance formula for difference between two variables)
a=(1/Age_Diff)
sd_slope=sqrt((a*sd_m1)^2+(a*sd_m2)^2-2*a*a*(sd_m1*sd_m2*m_corr))

#Cov(m1,slope)=cov(m1, a*(m2-m1)) =a*E[(m1-E[m1])((m2-m1)-E[m2-m1])]
#   =a*E[(m1-E[m1])(m2-E[m2])]- a* E[(m1-E[m1])(m1-E[m1])]
##   =a*Cov[m1,m2] - a*Var[m1]
##  =a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1
cov_V1=a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1

# Set covariance matrix
V1[1,1]=sd_m1^2
V1[2,2]=sd_slope^2
V1[1,2]=cov_V1
V1[2,1]=V1[1,2]
print(V1)
```

```{r Run Logit,results = "hide",echo = FALSE}
Menarche_Model_Data=data.frame(Age=menarche2$Age,Total=menarche2$Total,Menarche=menarche2$Menarche,Age2)
prior1=list(mu=mu1,Sigma=V1)
#glmb.out1<-glmb(n=10000,cbind(Menarche, Total-Menarche) ~ #Age2,family=binomial(logit),mu=mu1,Sigma=V1,data=Menarche_Model_Data)
glmb.out1<-glmb(n=10000,cbind(Menarche, Total-Menarche) ~ Age2,family=binomial(logit),prior=prior1,data=Menarche_Model_Data)

```


# General Discussion 

In this section, we will be continuing our discussion of the Bayesian model using the Menarche data that we covered in chapter 4.  Recall that our main model (with the carefully specified prior) had the summary output below. The focus of this chapter will be to dig deeper and to examine model fit, residuals, outliers, and the DIC statistic (for model comparisons).  

```{r Summary Logit}
summary(glmb.out1)
```

# Reviewing Model Fits

# Reviewing Model Residuals

# Using the DIC Statistic

# References
