---
title: "Chapter 4: Reviewing Model Predictions, Deviance Residuals and Model Statistics"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 4: Reviewing Model Predictions, Deviance Residuals and Model Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---
references:
- id: nygren2006
  title: Likelihood Subgradient Densities
  author:
  - family: Nygren
    given: Kjell
  - family: Nygren
    given: Lan Ma
  container-title: Journal of the American Statistical Association
  volume: 101
  URL: 'https://doi.org/10.1198/016214506000000357'
  DOI: 10.1198/016214506000000357
  issue: 475
  publisher: Taylor and Francis, Ltd
  page: 1144-1156
  type: article-journal
  issued:
    year: 2006
    month: 9

- id: Venables2002
  title: Modern Applied Statistics with S.
  author:
  - family: Venables
    given: W. N.
  - family: Ripley
    given: B. D.
  URL: 'https://www.springer.com/gp/book/9780387954578'
  DOI: 10.1007/978-0-387-21706-2
  publisher: Springer-Verlag New York
  page: 1-498
  type: book
  issued:
    year: 2002
    
- id: Hamilton2004
  title: Lecture notes obstetrics and gynaecology
  author:
  - family: Hamilton-Fairley
    given: Diana
  URL: 'https://www.bookdepository.com/Lecture-Notes-Obstetrics-Gynaecology-Diana-Hamilton-Fairley/9781405120661'
  publisher: Blackwell Pub.
  type: book
  issued:
    year: 2004
    month: 11

- id: Evans2006
  title: Checking for Prior-Data Conflict
  author:
  - family: Evans
    given: Michael
  - family: Moshonov
    given: Hadas
  container-title: Bayesian Analysis
  volume: 1
  URL: 'http://dx.doi.org/10.1214/06-BA129'
  DOI: 10.1214/06-BA129
  issue: 4
  page: 893-914
  type: article-journal
  issued:
    year: 2006
    month: 

- id: Spiegelhalter2002
  title: Bayesian measures of model complexity and fit
  author:
  - family: Spiegelhalter
    given: David J.
  - family: Best
    given: Nicola G.
  - family: Carlin
    given: Bradley P.
  - family: Van Der Linde
    given: Angelika
  container-title: Journal of the Royal Statistical Society Series B (Statistical Methodology)
  volume: 64
  URL: 'https://doi.org/10.1111/1467-9868.00353'
  DOI: 10.1111/1467-9868.00353
  issue: 4
  page: 583-639
  type: article-journal
  issued:
    year: 2002
    month: 9

    
---

---
nocite: | 
  @nygren2006
...



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(glmbayes)
```


```{r menarche data,results = "hide",echo = FALSE}
## Load menarche data
data(menarche2)
head(menarche2, 5)

```

```{r Analysis Setup,echo = FALSE}
## Number of variables in model
Age=menarche2$Age
nvars=2
set.seed(333)

## Reference Ages for setting of priors and Age_Difference
ref_age1=13  # user can modify this
ref_age2=15  ## user can modify this

## Define variables used later in analysis
Age2=Age-ref_age1
Age_Diff=ref_age2-ref_age1

```


```{r Prior Info,echo = FALSE}

## Point estimates at reference ages
m1=0.5  
m2=0.9

## Lower bound of prior credible intervals for point estimates
m1_lower=0.3
m2_lower=0.7

## Assumed correlation between the two (on link scale)
m_corr=0.4
```


```{r Logit: set up link function info and initialize prior matrices,echo = FALSE}

## Set up link function and initialize prior mean and Variance-Covariance matrices
bi_logit <- binomial(link="logit")
mu1<-matrix(0,nrow=nvars,ncol=1)
rownames(mu1)=c("Intercept","Age2")
colnames(mu1)=c("Prior Mean")
V1<-1*diag(nvars)
rownames(V1)=c("Intercept","Age2")
colnames(V1)=c("Intercept","Age2")
```


```{r Logit:set prior means,results = "hide",echo = FALSE}
## Prior mean for intercept is set to point estimate 
## at reference age1 (on logit scale)
mu1[1,1]=bi_logit$linkfun(m1)

## Prior mean for slope is set to difference in point estimates
## on logit scale divided by Age_Diff

mu1[2,1]=(bi_logit$linkfun(m2) -bi_logit$linkfun(m1))/Age_Diff 
print(mu1)

```

```{r Logit:set prior Variance Covariance matrix}
## Implied standard deviations for point estimates on logit scale

sd_m1= (bi_logit$linkfun(m1) -bi_logit$linkfun(m1_lower))/1.96
sd_m2= (bi_logit$linkfun(m2) -bi_logit$linkfun(m2_lower))/1.96

## Also compute implied estimate for upper bound of confidence intervals

m1_upper=bi_logit$linkinv(bi_logit$linkfun(m1)+sd_m1*1.96)
m2_upper=bi_logit$linkinv(bi_logit$linkfun(m2)+sd_m2*1.96)
print("m1_upper is:")
m1_upper
print("m2_upper is:")
m2_upper


## Implied Standard deviation for slope (using variance formula for difference between two variables)
a=(1/Age_Diff)
sd_slope=sqrt((a*sd_m1)^2+(a*sd_m2)^2-2*a*a*(sd_m1*sd_m2*m_corr))

#Cov(m1,slope)=cov(m1, a*(m2-m1)) =a*E[(m1-E[m1])((m2-m1)-E[m2-m1])]
#   =a*E[(m1-E[m1])(m2-E[m2])]- a* E[(m1-E[m1])(m1-E[m1])]
##   =a*Cov[m1,m2] - a*Var[m1]
##  =a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1
cov_V1=a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1

# Set covariance matrix
V1[1,1]=sd_m1^2
V1[2,2]=sd_slope^2
V1[1,2]=cov_V1
V1[2,1]=V1[1,2]
print("V1 is:")
print(V1)
```

```{r Run Logit,results = "hide",echo = FALSE}
Menarche_Model_Data=data.frame(Age=menarche2$Age,Total=menarche2$Total,Menarche=menarche2$Menarche,Age2)
prior1=list(mu=mu1,Sigma=V1)
#glmb.out1<-glmb(n=1000,cbind(Menarche, Total-Menarche) ~ #Age2,family=binomial(logit),mu=mu1,Sigma=V1,data=Menarche_Model_Data)
glmb.out1<-glmb(n=1000,cbind(Menarche, Total-Menarche) ~ Age2,family=binomial(logit),pfamily=dNormal(mu=mu1,Sigma=V1),data=Menarche_Model_Data)

```


# Introduction

In this section, we will be continuing our discussion of the Bayesian model using the Menarche data that we covered in chapter 4.  Recall that our main model (with the carefully specified prior) had the summary output below. The focus of this chapter will be to dig deeper and to examine model fit, residuals, outliers, and the DIC statistic (for model comparisons).  


# Reviewing Model Predictions

## Fits and Predictions for Actual Data


```{r Model_Preds1,fig.height = 5, fig.width = 7}
## Predicting for original data

pred1=predict(glmb.out1,type="response")
pred1_m=colMeans(pred1)

plot(Menarche/Total ~ Age, data=menarche2,main="% of girls Who have had their first period")
lines(menarche2$Age, pred1_m,lty=1)


```


## Predictions for Simulated Data

```{r Setup_NewData}
## Pull Model Info
mod_Object=glmb.out1
n=nrow(mod_Object$coefficients)
obs_size=median(menarche2$Total) ## Median of Total counts

# Generate New Data for The Independent Variable
Age_New <- seq(8, 20, 0.25)
Age2_New=Age_New-13

# olddata - Should contain all variables used in original model formula
olddata=data.frame(Age=menarche2$Age,
Menarche=menarche2$Menarche,Total=menarche2$Total,Age2=Age2)

# Should contain all variables used on RHS of model formula
newdata=data.frame(Age=Age_New,Age2=Age2_New)

```

```{r Model_Sims}
## Predict and Simulate for newdata

pred_menarche=predict(mod_Object,newdata=newdata,olddata=olddata,type="response")
pred_y=matrix(0,nrow=n,ncol=length(Age_New))
for(i in 1:n){
  pred_y[i,1:length(Age_New)]=rbinom(length(Age_New),size=obs_size,
  prob=pred_menarche[i,1:length(Age_New)])
}

```

```{r Mean_and_Quartile_Info}

## Produce mean and quantile information

pred_m=colMeans(pred_menarche)
pred_y_m=colMeans(pred_y/obs_size)
quant1_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.025))
quant2_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.975))
quant1_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.025))
quant2_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.975))

```


```{r Model_Plots,fig.height = 5, fig.width = 7}

plot(Menarche/Total ~ Age, data=menarche2,main="Percentage of girls Who have had their first period")
lines(Age_New, pred_m,lty=1)
lines(Age_New, quant1_m,lty=2)
lines(Age_New, quant2_m,lty=2)
lines(Age_New, quant1_m_y,lty=2)
lines(Age_New, quant2_m_y,lty=2)

```


```{r Store_Preds}

outdata=data.frame(Age=Age_New,Age2=Age2_New,pred_m=pred_m, quant1_m, quant2_m,quant1_m_y,quant2_m_y)

print("Prior Confidence Interval at age 13 was")
cbind(m1_lower,m1_upper)

print("Prior Confidence Interval at age 15 was")
cbind(m2_lower,m2_upper)

print("Posterior Predictions between ages 13 and 15 are")
outdata[which(outdata$Age >= 13&outdata$Age <= 15 ),]

```

# Reviewing Model Residuals

```{r Residual_Analysis}

## Get Residuals for model, their means, and credible bounds
res_out=residuals(glmb.out1)
res_mean=colMeans(res_out)
sqrt(mean(res_mean^2))

res_low1=apply(res_out,2,FUN=quantile,probs=c(0.025))
res_high1=apply(res_out,2,FUN=quantile,probs=c(0.975))
```

```{r Simulate_New_Residuals}
## Simulate new data and get residuals for simulated data

ysim1=simulate(glmb.out1,nsim=1,seed=NULL,pred=pred1,family="binomial",
                        prior.weights=weights(glmb.out1))
res_ysim_out1=residuals(glmb.out1,ysim=ysim1)
res_low=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.025))
res_high=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.975))
```


```{r Residual_Plot,fig.height = 5, fig.width = 7}
# Plot Credible Interval bounds for Deviance Residuals

plot(res_mean~Age,ylim=c(-2.5,2.5),
  main="Credible Interval Bounds for Logit Model Deviance Residuals",
  xlab = "Age", ylab = "Dev. Residuals")
lines(Age, 0*res_mean,lty=1)
lines(Age, res_low,lty=1)
lines(Age, res_high,lty=1)
lines(Age, res_low1,lty=2)
lines(Age, res_high1,lty=2)
```


```{r Raw_Data}
# Look a the raw data
menarche2
```

# Using the DIC Statistic

The DIC Statistic is mainly useful when comparing the predictions from several models with the same likelihood function but with different specifications (say different link-functions or different model variables). Here, we will explore alternative link functions for the Menarche data by using the Probit and Clog-Log Models and will then compare the models using the DIC Statistic and related outputs.

## Running The Probit Model

```{r Probit_Model}
## ----Probit: set up link function info and initialize prior matrices-----------

## Set up link function and initialize prior mean and Variance-Covariance matrices
bi_probit <- binomial(link="probit")
mu2<-matrix(0,nrow=nvars,ncol=1)
rownames(mu2)=c("Intercept","Age2")
colnames(mu2)=c("Prior Mean")
V2<-1*diag(nvars)
rownames(V2)=c("Intercept","Age2")
colnames(V2)=c("Intercept","Age2")


## ----Probit:set prior means----------------------------------------------------
## Prior mean for intercept is set to point estimate 
## at reference age1 (on logit scale)
mu2[1,1]=bi_probit$linkfun(m1)

## Prior mean for slope is set to difference in point estimates
## on logit scale divided by Age_Diff

mu2[2,1]=(bi_probit$linkfun(m2) -bi_probit$linkfun(m1))/Age_Diff 
print(mu2)


## ----Probit:set prior Variance Covariance matrix-------------------------------
## Implied standard deviations for point estimates on logit scale

sd_m1= (bi_probit$linkfun(m1) -bi_probit$linkfun(m1_lower))/1.96
sd_m2= (bi_probit$linkfun(m2) -bi_probit$linkfun(m2_lower))/1.96

## Implied Standard deviation for slope (using variance formula for difference between two variables)
a=(1/Age_Diff)
sd_slope=sqrt((a*sd_m1)^2+(a*sd_m2)^2-2*a*a*(sd_m1*sd_m2*m_corr))

#Cov(m1,slope)=cov(m1, a*(m2-m1)) =a*E[(m1-E[m1])((m2-m1)-E[m2-m1])]
#   =a*E[(m1-E[m1])(m2-E[m2])]- a* E[(m1-E[m1])(m1-E[m1])]
##   =a*Cov[m1,m2] - a*Var[m1]
##  =a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1
cov_V2=a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1

# Set covariance matrix
V2[1,1]=sd_m1^2
V2[2,2]=sd_slope^2
V2[1,2]=cov_V2
V2[2,1]=V2[1,2]
print(V2)


## ----Run Probit,results = "hide"-----------------------------------------------
Menarche_Model_Data=data.frame(Age=menarche2$Age,Total=menarche2$Total,
                               Menarche=menarche2$Menarche,Age2)

glmb.out2<-glmb(n=1000,cbind(Menarche, Total-Menarche) ~Age2,family=binomial(probit),
                pfamily=dNormal(mu=mu2,Sigma=V2),data=Menarche_Model_Data)

print(glmb.out2)
summary(glmb.out2)
```

## Running The Clog-Log Model


```{r cloglog_Model}

## ----cloglog: set up link function info and initialize prior matrices-----------

## Set up link function and initialize prior mean and Variance-Covariance matrices
bi_cloglog <- binomial(link="cloglog")
mu3<-matrix(0,nrow=nvars,ncol=1)
rownames(mu3)=c("Intercept","Age2")
colnames(mu3)=c("Prior Mean")
V3<-1*diag(nvars)
rownames(V3)=c("Intercept","Age2")
colnames(V3)=c("Intercept","Age2")


## ----cloglog:set prior means----------------------------------------------------
## Prior mean for intercept is set to point estimate 
## at reference age1 (on logit scale)
mu3[1,1]=bi_cloglog$linkfun(m1)

## Prior mean for slope is set to difference in point estimates
## on logit scale divided by Age_Diff

mu3[2,1]=(bi_cloglog$linkfun(m2) -bi_cloglog$linkfun(m1))/Age_Diff 
print(mu3)


## ----Probit:set prior Variance Covariance matrix-------------------------------
## Implied standard deviations for point estimates on logit scale

sd_m1= (bi_cloglog$linkfun(m1) -bi_cloglog$linkfun(m1_lower))/1.96
sd_m2= (bi_cloglog$linkfun(m2) -bi_cloglog$linkfun(m2_lower))/1.96

## Implied Standard deviation for slope (using variance formula for difference between two variables)
a=(1/Age_Diff)
sd_slope=sqrt((a*sd_m1)^2+(a*sd_m2)^2-2*a*a*(sd_m1*sd_m2*m_corr))

#Cov(m1,slope)=cov(m1, a*(m2-m1)) =a*E[(m1-E[m1])((m2-m1)-E[m2-m1])]
#   =a*E[(m1-E[m1])(m2-E[m2])]- a* E[(m1-E[m1])(m1-E[m1])]
##   =a*Cov[m1,m2] - a*Var[m1]
##  =a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1
cov_V3=a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1

# Set covariance matrix
V3[1,1]=sd_m1^2
V3[2,2]=sd_slope^2
V3[1,2]=cov_V3
V3[2,1]=V3[1,2]
print(V3)


## ----Run cloglog,results = "hide"-----------------------------------------------
Menarche_Model_Data=data.frame(Age=menarche2$Age,Total=menarche2$Total,
                               Menarche=menarche2$Menarche,Age2)

glmb.out3<-glmb(n=1000,cbind(Menarche, Total-Menarche) ~Age2,family=binomial(cloglog),
                pfamily=dNormal(mu=mu3,Sigma=V3),data=Menarche_Model_Data)

print(glmb.out3)
summary(glmb.out3)


```
## Running The Reverse Clog-Log Model


```{r cloglog_Model2}

## ----cloglog: set up link function info and initialize prior matrices-----------

## Set up link function and initialize prior mean and Variance-Covariance matrices
bi_cloglog <- binomial(link="cloglog")
mu4<-matrix(0,nrow=nvars,ncol=1)
rownames(mu4)=c("Intercept","Age2")
colnames(mu4)=c("Prior Mean")
V4<-1*diag(nvars)
rownames(V4)=c("Intercept","Age2")
colnames(V4)=c("Intercept","Age2")


## ----cloglog:set prior means----------------------------------------------------
## Prior mean for intercept is set to point estimate 
## at reference age1 (on logit scale)
mu4[1,1]=bi_cloglog$linkfun((1-m1))

## Prior mean for slope is set to difference in point estimates
## on logit scale divided by Age_Diff

mu4[2,1]=(bi_cloglog$linkfun((1-m2))-bi_cloglog$linkfun((1-m1)))/Age_Diff 
print(mu4)


## ----Probit:set prior Variance Covariance matrix-------------------------------
## Implied standard deviations for point estimates on logit scale

sd_m1= (bi_cloglog$linkfun((1-m1_lower))-bi_cloglog$linkfun((1-m1)))/1.96
sd_m2= (bi_cloglog$linkfun((1-m2_lower))-bi_cloglog$linkfun((1-m2)))/1.96

## Implied Standard deviation for slope (using variance formula for difference between two variables)
a=(1/Age_Diff)
sd_slope=sqrt((a*sd_m1)^2+(a*sd_m2)^2-2*a*a*(sd_m1*sd_m2*m_corr))

#Cov(m1,slope)=cov(m1, a*(m2-m1)) =a*E[(m1-E[m1])((m2-m1)-E[m2-m1])]
#   =a*E[(m1-E[m1])(m2-E[m2])]- a* E[(m1-E[m1])(m1-E[m1])]
##   =a*Cov[m1,m2] - a*Var[m1]
##  =a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1
cov_V4=a*sd_m1*sd_m2*m_corr-a* sd_m1*sd_m1

# Set covariance matrix
V4[1,1]=sd_m1^2
V4[2,2]=sd_slope^2
V4[1,2]=cov_V4
V4[2,1]=V4[1,2]
print(V4)


## ----Run cloglog,results = "hide"-----------------------------------------------
Menarche_Model_Data=data.frame(Age=menarche2$Age,Total=menarche2$Total,
                               Menarche=menarche2$Menarche,Age2)

glmb.out4<-glmb(n=1000,cbind(Total-Menarche,Menarche) ~Age2,family=binomial(cloglog),
                pfamily=dNormal(mu=mu4,Sigma=V4),data=Menarche_Model_Data)

print(glmb.out4)
summary(glmb.out4)

```

## Comparing the DIC Statistics

```{r Compare_DIC}
DIC_Out=rbind(
  extractAIC(glmb.out1),
  extractAIC(glmb.out2),
  extractAIC(glmb.out3),
  extractAIC(glmb.out4))

rownames(DIC_Out)=c("Logit","Probit","Clog-Log","Reverse Clog-Log")
DIC_Out
```

```{r Compare_AIC}

glm.out1=glm(cbind(Menarche, Total-Menarche) ~Age2,family=binomial(logit),data=Menarche_Model_Data)
glm.out2=glm(cbind(Menarche, Total-Menarche) ~Age2,family=binomial(probit),data=Menarche_Model_Data)
glm.out3=glm(cbind(Menarche, Total-Menarche) ~Age2,family=binomial(cloglog),data=Menarche_Model_Data)
glm.out4=glm(cbind(Total-Menarche,Menarche ) ~Age2,family=binomial(cloglog),data=Menarche_Model_Data)

AIC_Out=rbind(extractAIC(glm.out1),
extractAIC(glm.out2),
extractAIC(glm.out3),
extractAIC(glm.out4))

rownames(AIC_Out)=c("Logit","Probit","Clog-Log","Reverse Clog-Log")
colnames(AIC_Out)=c("edf","AIC")
AIC_Out
```

## Plotting Predictions and Residuals For the Additional Models

### Probit Model

*Predictions*
```{r Model_Sims2}
## Predict and Simulate for newdata

mod_Object=glmb.out2
pred1=predict(glmb.out2,type="response")
pred1_m=colMeans(pred1)
n=nrow(mod_Object$coefficients)
pred_menarche=predict(mod_Object,newdata=newdata,olddata=olddata,type="response")
pred_y=matrix(0,nrow=n,ncol=length(Age_New))
for(i in 1:n){
  pred_y[i,1:length(Age_New)]=rbinom(length(Age_New),size=obs_size,
  prob=pred_menarche[i,1:length(Age_New)])
}

```

```{r Mean_and_Quartile_Info2}

## Produce mean and quantile information

pred_m=colMeans(pred_menarche)
pred_y_m=colMeans(pred_y/obs_size)
quant1_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.025))
quant2_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.975))
quant1_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.025))
quant2_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.975))

```


```{r Model_Plots2,fig.height = 5, fig.width = 7}

plot(Menarche/Total ~ Age, data=menarche2,main="Percentage of girls Who have had their first period")
lines(Age_New, pred_m,lty=1)
lines(Age_New, quant1_m,lty=2)
lines(Age_New, quant2_m,lty=2)
lines(Age_New, quant1_m_y,lty=2)
lines(Age_New, quant2_m_y,lty=2)

```
*Residuals*
```{r Residual_Analysis2}
## Get Residuals for model, their means, and credible bounds
res_out=residuals(glmb.out2)
res_mean=colMeans(res_out)
sqrt(mean(res_mean^2))
res_low1=apply(res_out,2,FUN=quantile,probs=c(0.025))
res_high1=apply(res_out,2,FUN=quantile,probs=c(0.975))
```

```{r Simulate_New_Residuals2}
## Simulate new data and get residuals for simulated data

ysim1=simulate(glmb.out2,nsim=1,seed=NULL,pred=pred1,family="binomial",
                        prior.weights=weights(glmb.out2))
res_ysim_out1=residuals(glmb.out2,ysim=ysim1)
res_low=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.025))
res_high=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.975))
```


```{r Residual_Plot2,fig.height = 5, fig.width = 7}
# Plot Credible Interval bounds for Deviance Residuals

plot(res_mean~Age,ylim=c(-2.5,2.5),
  main="Credible Interval Bounds for Probit Model Deviance Residuals",
  xlab = "Age", ylab = "Dev. Residuals")
lines(Age, 0*res_mean,lty=1)
lines(Age, res_low,lty=1)
lines(Age, res_high,lty=1)
lines(Age, res_low1,lty=2)
lines(Age, res_high1,lty=2)
```

### clog-log Model

*Predictions*
```{r Model_Sims3}
## Predict and Simulate for newdata
mod_Object=glmb.out3
pred1=predict(glmb.out3,type="response")
pred1_m=colMeans(pred1)
n=nrow(mod_Object$coefficients)
pred_menarche=predict(mod_Object,newdata=newdata,olddata=olddata,type="response")
pred_y=matrix(0,nrow=n,ncol=length(Age_New))
for(i in 1:n){
  pred_y[i,1:length(Age_New)]=rbinom(length(Age_New),size=obs_size,
  prob=pred_menarche[i,1:length(Age_New)])
}

```

```{r Mean_and_Quartile_Info3}

## Produce mean and quantile information

pred_m=colMeans(pred_menarche)
pred_y_m=colMeans(pred_y/obs_size)
quant1_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.025))
quant2_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.975))
quant1_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.025))
quant2_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.975))

```


```{r Model_Plots3,fig.height = 5, fig.width = 7}

plot(Menarche/Total ~ Age, data=menarche2,main="Percentage of girls Who have had their first period")
lines(Age_New, pred_m,lty=1)
lines(Age_New, quant1_m,lty=2)
lines(Age_New, quant2_m,lty=2)
lines(Age_New, quant1_m_y,lty=2)
lines(Age_New, quant2_m_y,lty=2)

```
*Residuals*
```{r Residual_Analysis3}
## Get Residuals for model, their means, and credible bounds
res_out=residuals(glmb.out3)
res_mean=colMeans(res_out)
sqrt(mean(res_mean^2))
res_low1=apply(res_out,2,FUN=quantile,probs=c(0.025))
res_high1=apply(res_out,2,FUN=quantile,probs=c(0.975))
```

```{r Simulate_New_Residuals3}
## Simulate new data and get residuals for simulated data

ysim1=simulate(glmb.out3,nsim=1,seed=NULL,pred=pred1,family="binomial",
                        prior.weights=weights(glmb.out3))
res_ysim_out1=residuals(glmb.out3,ysim=ysim1)
res_low=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.025))
res_high=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.975))
```


```{r Residual_Plot3,fig.height = 5, fig.width = 7}
# Plot Credible Interval bounds for Deviance Residuals

plot(res_mean~Age,ylim=c(-4.0,4.0),
  main="Credible Interval Bounds for cloglog Model Deviance Residuals",
  xlab = "Age", ylab = "Dev. Residuals")
lines(Age, 0*res_mean,lty=1)
lines(Age, res_low,lty=1)
lines(Age, res_high,lty=1)
lines(Age, res_low1,lty=2)
lines(Age, res_high1,lty=2)
```

### Reverse clog-log Model

*Predictions*
```{r Model_Sims4}
## Predict and Simulate for newdata
mod_Object=glmb.out4
pred1=predict(glmb.out4,type="response")
pred1_m=colMeans(pred1)
n=nrow(mod_Object$coefficients)
pred_menarche=predict(mod_Object,newdata=newdata,olddata=olddata,type="response")
pred_y=matrix(0,nrow=n,ncol=length(Age_New))
for(i in 1:n){
  pred_y[i,1:length(Age_New)]=rbinom(length(Age_New),size=obs_size,
  prob=pred_menarche[i,1:length(Age_New)])
}

```

```{r Mean_and_Quartile_Info4}

## Produce mean and quantile information

pred_m=colMeans(pred_menarche)
pred_y_m=colMeans(pred_y/obs_size)
quant1_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.025))
quant2_m=apply(pred_menarche,2,FUN=quantile,probs=c(0.975))
quant1_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.025))
quant2_m_y=apply(pred_y/obs_size,2,FUN=quantile,probs=c(0.975))

```


```{r Model_Plots4,fig.height = 5, fig.width = 7}

plot(Menarche/Total ~ Age, data=menarche2,main="Percentage of girls Who have had their first period")
lines(Age_New, 1-pred_m,lty=1)
lines(Age_New, 1-quant1_m,lty=2)
lines(Age_New, 1-quant2_m,lty=2)
lines(Age_New, 1-quant1_m_y,lty=2)
lines(Age_New, 1-quant2_m_y,lty=2)

```
*Residuals*
```{r Residual_Analysis4}
## Get Residuals for model, their means, and credible bounds
res_out=residuals(glmb.out4)
res_mean=colMeans(res_out)
sqrt(mean(res_mean^2))
res_low1=apply(res_out,2,FUN=quantile,probs=c(0.025))
res_high1=apply(res_out,2,FUN=quantile,probs=c(0.975))
```

```{r Simulate_New_Residuals4}
## Simulate new data and get residuals for simulated data

ysim1=simulate(glmb.out4,nsim=1,seed=NULL,pred=pred1,family="binomial",
                        prior.weights=weights(glmb.out4))
res_ysim_out1=residuals(glmb.out4,ysim=ysim1)
res_low=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.025))
res_high=apply(res_ysim_out1,2,FUN=quantile,probs=c(0.975))
```


```{r Residual_Plot4,fig.height = 5, fig.width = 7}
# Plot Credible Interval bounds for Deviance Residuals

plot(-res_mean~Age,ylim=c(-3.0,3.0),
  main="Credible Interval Bounds - Rev. clog-log Model Deviance Residuals",
  xlab = "Age", ylab = "Dev. Residuals")
lines(Age, 0*res_mean,lty=1)
lines(Age, -res_low,lty=1)
lines(Age, -res_high,lty=1)
lines(Age, -res_low1,lty=2)
lines(Age, -res_high1,lty=2)
```


# References
